---
title: "ISE Analysis"
author: "J. Beem-Miller"
date: "17 April 2020"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document:
    latex_engine: xelatex
---

# Introduction
This notebook contains code for reading radiocarbon data from the Ice Storm Experiment (ISE). The ISE project was initiated by Prof. Tim Fahey at Cornell University at Arnot Forest, a northern hardwood research forest maintained by Cornell University near the main campus in upstate New York. Geographic coordinates for the site are: 42.26, -76.627 (obtained from google maps; these are not coordinates for the plots themselves). 

# Background
*The following notes are drawn from information supplied by T. Fahey in the file:* 
"ise-fahey/data/external/ise_fahey_plot-IDs_2018-12-29/README.yaml"

Fahey and collaborators simulated the effect of an ice storm on a series of plots by spraying the canopy of mature forest plots with water during mid-winter of 2016. The treatments successfully damaged the canopies of the trees. The design included two replicate plots for each of four treatments: low, mid, high and mid 2x designating different levels of ice. “2x” plots were treated in two successive years, 2016 and 2017. 

The hypothesis for the experiment is that canopy damage and subsequent recovery would result in reduced C supply and increased C demand, so that trees would need to draw on older C reserves to build fine roots following the treatment. 

Data analyzed in this notebook come from fine roots sampled from ingrowth cores;  the codes “O” and “M” are used for organic samples from the forest floor and samples from mineral soil, respectively. Roots analyzed were sampled in the second year of the experiment (2017). Nine ingrowth cores were collected per plot, which were pooled in the lab and then split into three lab replicates for analysis. Only data from the "high", "mid2x", and "control" treatments were analyzed.
```{r setup, include = FALSE}
library(ggplot2)
library(dplyr)
library(lme4)
library(car)
```

```{r read-data, echo = FALSE}
# read data (tab 'R_ISE_dd-mm-yyyy' from "Fahey_ISE_FineRoots_15-05-2019.xlsx")
ise <- read.csv("../data/derived/2020-04-17/ise_14c-root_2020-04-17.csv")
# create new IDs from treat, horizon, rep
ise$ID <- paste(ise$Treat, ise$Horizon, ise$Rep, sep = "_")
# combine UCI and Jena 14c data
ise$d14C <- ifelse(is.na(ise$d14_UCI), ise$d14_JEN, ise$d14_UCI)
# factorize Plot
ise$plot_f <- factor(ise$Plot)
```
# Results
Plotting the raw data by treatment ("Ctl" = control; "high" = high, treated in 2016 only; "mid2x" = mid level treatment applied in both 2016 and 2017) shows a potential treatment response in the mineral horizon samples (horizon = "M"), but no clear effect in the organic horizon samples. Overall, variability is high (Fig. 1, below).
```{r plot-14c, echo = FALSE}
ggplot(ise, aes(Treat, d14C, color = Horizon, shape = plot_f)) +
  geom_point(size = 2) +
  facet_grid(rows = vars(Horizon)) +
  ggtitle(expression('Fig 1. Fine root' ~ Delta*''^14*'C (\u2030)')) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold"))
```

Figures 2a and 2b show data aggregated by treatment (below). However, the different response in the different plots suggests that combining data from different plots may obscure significant plot-specific trends, e.g. mineral horizon samples from plot 1 vs. plot 8 for the "mid2x" treatment, and weight outliers too heavily (e.g.)
```{r aggregate data, echo = FALSE}
ise.avg <- data.frame(Treat = rep(sort(unique(ise$Treat)),2),
                      Horizon = rep(unique(ise$Horizon),each=3))
ise.avg$d14C.avg <- c(unlist(tapply(ise[ise$Horizon=="M","d14C"], ise[ise$Horizon=="M","Treat"], mean)),
                      unlist(tapply(ise[ise$Horizon=="O","d14C"], ise[ise$Horizon=="O","Treat"], mean)))
ise.avg$d14C.sd <- c(unlist(tapply(ise[ise$Horizon=="M","d14C"], ise[ise$Horizon=="M","Treat"], sd)),
                     unlist(tapply(ise[ise$Horizon=="O","d14C"], ise[ise$Horizon=="O","Treat"], sd)))
ise.avg$d14C.se <- ise.avg$d14C.sd/sqrt(6)
ise.avg$u <- ise.avg$d14C.avg + 1.96*ise.avg$d14C.se
ise.avg$l <- ise.avg$d14C.avg - 1.96*ise.avg$d14C.se
```
```{r plot-14c w/ error bars, echo = FALSE}
# plot w/ 95% CI
ggplot(ise.avg, aes(Treat, d14C.avg, fill=Horizon)) +
  geom_col(position = "dodge", width=.9) +
  geom_errorbar(aes(ymin=l, ymax=u, group=Horizon), position=position_dodge(.9), width=.5) +
  facet_grid(rows = vars(Horizon)) +
  ggtitle(expression('Fig 2a. Fine root' ~ Delta*''^14*'C (\u2030, \u00B1 SE)')) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold"))
# plot w/ se
ggplot(ise.avg, aes(Treat, d14C.avg, fill=Horizon)) +
  geom_col(position = "dodge", width=.9) +
  geom_errorbar(aes(ymin=d14C.avg-d14C.se, ymax=d14C.avg+d14C.se, group=Horizon), position=position_dodge(.9), width=.5) +
  facet_grid(rows = vars(Horizon)) +
  ggtitle(expression('Fig 2b. Fine root' ~ Delta*''^14*'C (\u2030, \u00B1 95% CI)')) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold"))
```

To assess the effect of individual plots, first we will create a variable representing the difference in $\Delta$^14^C between treatment and control samples, aggregated by plot.

```{r aggregate by plot, echo = FALSE, include = FALSE}
# calculate mean control d14C
ise.ctl <- filter(ise, Treat == "Ctl") %>%
  group_by(Horizon) %>%
  summarise(mean = mean(d14C, na.rm = TRUE),
            sd = sd(d14C, na.rm = TRUE),
            count = n()) %>%
  mutate(se = sd / sqrt(count),
         ci95 = 1.96 * sd / sqrt(count))
CI95.ctl.m <- unlist(ise.ctl[ise.ctl$Horizon == "M", "ci95"])
CI95.ctl.o <- unlist(ise.ctl[ise.ctl$Horizon == "O", "ci95"])

# calculate dif in 14C from control for M, O samples
ise$dif <- ifelse(ise$Horizon == "M", 
                  ise$d14C-CI95.ctl.m, 
                  ise$d14C-CI95.ctl.o)
ise.plot <- filter(ise, Treat != "Ctl") %>%
  group_by(Horizon, plot_f) %>%
  summarise(mean = mean(dif),
            sd = sd(dif),
            count = n()) %>%
  mutate(se = sd / sqrt(count),
         l = mean - 1.96 * se,
         u = mean + 1.96 * se)
ise.plot$Treat <- ise[match(ise.plot$plot_f, ise$plot_f), "Treat"]
```

Plotting the plot-level differences from the mean of the controls, by horizon, makes the differences appear more significant (Fig. 3a and 3b). However, with only three replicates per plot, perhaps these differences are not truly significant. 
```{r plot-aggregated plots, echo = FALSE}
# plot min
ise.plot %>% filter(Horizon == "M") %>%
  ggplot(., aes(plot_f, mean)) +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = 1.79, linetype = "dashed") +
  geom_hline(yintercept = -1.79, linetype = "dashed") +
  geom_point(size = 3, color = "#F8766D") +
  geom_errorbar(aes(ymin = l, 
                    ymax = u, 
                    group = Horizon),
                width=.2,
                color = "#F8766D") +
  facet_grid(cols = vars(Treat)) +
  ggtitle(expression('Fig 3a. Mineral soil fine root' ~ Delta*''^14*'C by plot (\u2030, \u00B1 95% CI)')) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold"))

# plot org
ise.plot %>% filter(Horizon == "O") %>%
  ggplot(., aes(plot_f, mean, drop = TRUE)) +
  geom_hline(yintercept = 0) +
  geom_hline(yintercept = 3.91, linetype = "dashed") +
  geom_hline(yintercept = -3.91, linetype = "dashed") +
  geom_point(size = 3, color = "#00BFC4") +
  geom_errorbar(aes(ymin = l, 
                    ymax = u, 
                    group = Horizon),
                width=.2,
                color = "#00BFC4") +
  facet_grid(cols = vars(Treat)) +
  ggtitle(expression('Fig 3b. Organic soil fine root' ~ Delta*''^14*'C by plot (\u2030, \u00B1 95% CI)')) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold"))
```

Statistically, treatment differences are not signficant at alpha = 0.05 when analyzed in a linear regression model: neither for the mineral ("M") or organic ("O") horizon samples.
```{r OLS, echo = FALSE}
# Linear regression of d14C by treatment, by horizon
summary(lm(d14C ~ Treat, ise[ise$Horizon == "M", ])) # marginal signficance of "high" treatment
summary(lm(d14C ~ Treat, ise[ise$Horizon == "O", ])) # ns
```

Contrasts are also not significant:
```{r TukeyHSD, echo = FALSE}
# Contrasts for mineral samples:
TukeyHSD(aov(d14C ~ Treat, ise[ise$Horizon == "M",]))

# Contrasts for organic samples:
TukeyHSD(aov(d14C ~ Treat, ise[ise$Horizon == "O",]))
```

A mixed model approach with plot considered as a random factor shows the same results: while there is no signficant treatment effect for organic horizon samples, treatment effects are marginally signficant (Pr(>Chisq)) for mineral samples, which we know is heavily influenced by the data from rep 1 of the "high" damage treatment (sample ID = high_M_1).
```{r linear mixed model, echo = TRUE}
## Mixed model, considering effect of plot
# Min samples
ise.mm.m <- lmer(d14C ~ Treat + (1 | plot_f), REML = FALSE, data = ise[ise$Horizon == "M", ])
summary(ise.mm.m)
# compare unrestricted to restricted fit, i.e. w/o effect of Treat
ise.mm.m.0 <- lmer(d14C ~ (1 | plot_f), REML = F, ise[ise$Horizon == "M", ])
anova(ise.mm.m.0, ise.mm.m)

# Org samples:
ise.mm.o <- lmer(d14C ~ Treat + (1 | plot_f), data = ise[ise$Horizon == "O", ], REML = FALSE)
summary(ise.mm.o)
# compare unrestricted to restricted fit, i.e. w/o effect of Treat
ise.mm.o.0 <- lmer(d14C ~ (1 | plot_f), REML = F, ise[ise$Horizon == "O", ])
anova(ise.mm.o.0, ise.mm.o)

# Wald test for significance of Treatment in mixed models
Anova(ise.mm.m) # marginal signficance
Anova(ise.mm.o) # ns
```

# Conclusion
Overall, there is no evidence for treatment effects in the organic horizon samples, but some evidence of enrichment in mineral soil samples in response to treatment---especially the "high" damage treatment. The observed effect is in line with the hypothesis, i.e. mobilization of older (more enriched) carbon reserves in response to damage. 